---
title: "U.S. Electric Grid: Will it power the future?"
author: "Ben Mayes"
date: "2025-06-01"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Install and Load Required Libraries
```{r}
options(repos=c(CRAN="https://cran.r-project.org"))
install.packages("forecast")
install.packages("readr")
install.packages("stats")
install.packages("tseries")
install.packages("lmtest")
library(forecast)
library(readr)
library(forecast)
library(stats)
library(tseries)
library(lmtest)
```
# Read in Demand Data
```{r}
data <- read.csv("C:/Users/00815/OneDrive/Desktop/MSU/CIS 663/CIS 663 Project/Demand_per_month.csv")
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```
# Read in Generation Data
```{r}
data2 <- read.csv("C:/Users/00815/OneDrive/Desktop/MSU/CIS 663/CIS 663 Project/Generation_per_month.csv")
data2$date <- as.Date(data2$date, format = "%Y-%m-%d")
```
# View Data Head
```{r}
head(data)
head(data2)
```
# Create time series for demand and generation
```{r}
data_ts <- ts(data$demand, start = c(2010, 1), frequency = 12)
data2_ts <- ts(data2$generation, start = c(2010, 1), frequency = 12)
```

# Plot demand and generation time series
```{r}
plot(data_ts, main = "U.S. Electric Grid Generation and Demand", ylab = "Generation(red) / Demand(blue) (TWh)", xlab = "Year", col = "blue", lwd = 2)
lines(data2_ts, col = "red", lwd = 2)
```

# View summary of time series
```{r}
summary(data_ts)
summary(data2_ts)
```

# Acf and Pacf plots for demand and generation
```{r}
par(mfrow = c(2, 2))
acf(data_ts, main = "ACF of Demand")
acf(data2_ts, main = "ACF of Generation")
pacf(data_ts, main = "PACF of Demand")
pacf(data2_ts, main = "PACF of Generation")
```

# Check for stationarity
```{r}
adf_test <- adf.test(data_ts)
if (adf_test$p.value < 0.05) {
  print("The time series is stationary.")
} else {
  print("The time series is not stationary. Differencing may be required.")
}


adf_test2 <- adf.test(data2_ts)
if (adf_test2$p.value < 0.05) {
  print("The time series is stationary.")
} else {
  print("The time series is not stationary. Differencing may be required.")
}
```

# Train and test split demand data
```{r}
train_size <- floor(0.8 * length(data_ts))
train_data <- window(data_ts, end = c(2019, 12))
test_data <- window(data_ts, start = c(2020, 1))
```

# Summary of the training and test data
```{r}
summary(train_data)
summary(test_data)
```

# Fit ARIMA model
```{r}
fit <- auto.arima(train_data)
```

# View Arima model summary
```{r}
summary(fit)
```

# Check residuals of the fitted model demand
```{r}
checkresiduals(fit)
```

# Forecast using the fitted model for the next 60 months
```{r}
forecasted_values <- forecast(fit, h = 120)
```

# View the forecasted values
```{r}
print(forecasted_values)
```

# Plot the forecast values against the actual test data demand
```{r}
plot(forecasted_values, main = "ARIMA Forecast vs Actual Demand", ylab = "Demand Forecast (TWh)", xlab = "Year", col = "blue", lwd = 2)
lines(test_data, col = "red", lwd = 2)
```

#Train and fit generation data
```{r}
train_size2 <- floor(0.8 * length(data2_ts))
train_data2 <- window(data2_ts, end = c(2019, 12))
test_data2 <- window(data2_ts, start = c(2020, 1))
```

# Summary of the training and test data for generation
```{r}
summary(train_data2)
summary(test_data2)
```

# Fit ARIMA model for generation data
```{r}
fit2 <- auto.arima(train_data2)
```
# View ARIMA model summary for generation data
```{r}
summary(fit2)
```
# Check residuals of the fitted model for generation data
```{r}
checkresiduals(fit2)
```

# Forecast using the fitted model for generation data
```{r}
forecasted_values2 <- forecast(fit2, h = 120)
```

# Plot the forecasted values against the actual test data for generation
```{r}
plot(forecasted_values2, main = "ARIMA Forecast vs Actual Generation", ylab = "Generation Forecast (Twh)", xlab = "Year", col = "blue", lwd = 2)
lines(test_data2, col = "red", lwd = 2)
```

# Summary of forecasted_values$mean
```{r}
print(summary(forecasted_values$mean))
```
# Summary of test_data
```{r}
print(summary(test_data))
```

# T-test of forecasted values against actual values demand
```{r}
t_test_demand <- t.test(forecasted_values$mean, test_data)
if (t_test_demand$p.value < 0.05) {
  print("The forecasted demand values are significantly different from the actual values.")
} else {
  print("The forecasted demand values are not significantly different from the actual values.")
}
```

# Results of t-test
```{r}
print(t_test_demand)
```

# T-test of forecasted values against actual values for generation
```{r}
t_test_generation <- t.test(forecasted_values2$mean, test_data2)
if (t_test_generation$p.value < 0.05) {
  print("The forecasted generation values are significantly different from the actual values.")
} else {
  print("The forecasted generation values are not significantly different from the actual values.")
}
```

# Results of t-test
```{r}
print(t_test_generation)
```


# T-Test if predicted values of generation are significantly different from demand
```{r}
t_test_comparison <- t.test(forecasted_values$mean, forecasted_values2$mean)
if (t_test_comparison$p.value < 0.05) {
  print("The forecasted demand and generation values are significantly different.")
} else {
  print("The forecasted demand and generation values are not significantly different.")
}
```

# Results of t-test
```{r}
print(t_test_comparison)
```



